A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Additional Methods

### 2020-01-05 writeUp/supplementary_materials_methods.tex
updated SOM and MS to include higher kappa estimates

\begin{itemize}
	\item 69 and 70: Since the release was not disaggregated by sex, the recapture data were aggregated across both males and females, resulting in a single time series in each case.
	\item 482: There are actually two series (one with captured mosquitoes, the other with young reared) in the paper \cite[]{rawlings1981influence}, whereas the database contains only one recapture series. In this analysis, both time series are included.
	\item 483: There are three studies in the paper \cite[]{rawlings1981influence}, (by dye colour there is one for `M\&Y', another for white, and a third for green), whereas the database only contains one release series. In this analysis, we analyse all three time series.
	\item 578: A null observation was removed on day 12, since isn't present in the original paper \cite[]{eyles1943experiment}.
	\item 585: It is not clear from the original paper \cite[]{smith1941observations} whether the missing recapture observations on certain days are because no mosquitoes were found on those days, or because no efforts were made to recapture them on those days. We have assumed that the number recaptured were as stated in the database on those days (missing, because no recapture efforts were made on those days).
	\item 586: It was unclear from database whether the missing observations are due to lack of recapture, or no effort. Have assumed that zero mosquitoes were captured on these days as seems most likely (the paper could not be accessed).
	\item 342: Changed number of released from 6,000 to 12,000 following the original publication \cite[]{bryan1991anopheles}.
	\item 246: Moved the observations from captures to recaptures following the original publication \cite[]{arredondo1998gonotrophic}.
	\item 55: Corrected a mistake in release data; the actual number released is 739, not 749 \cite[]{midega2007estimating}.
	\item 478-481: All four of these series give releases not disaggregated by gender, whereas the recaptures are split this way. Have combined the recapture series to yield a single release-recapture series in each case.
	\item 154: Moved captured male series to recaptured following original paper \cite[]{tsuda2001movement}. 
	\item 334: Data starts on day 3, thus have added this number of days to our time variable in this case.
\end{itemize}

The R file \verb|s_clean_data.R| attached cleans the data, makes the above amendments and outputs the data in a form amenable for estimation using Stan \citep{carpenter2016stan}.

@@

\begin{table}[htbp]
	\centering
	\begin{tabular}{lc}
		\toprule
		\textbf{Data characteristic} & \textbf{Count} \\
		\midrule
		Species & 33 \\
		Genera & 3 \\
		\textit{Anopheles} & 94 \\
		\textit{Aedes} & 91 \\
		\textit{Culex} & 47 \\
		Female-only MRR & 179 \\
		Male-only MRR & 35 \\
		Mixed sex MRR & 18 \\
		Pre-release blood-feeding only MRR & 71 \\
		Pre-release sugar-feeding only MRR & 41 \\
		Pre-release both blood- and sugar-feeding MRR & 4 \\
		Pre-release neither blood- and sugar-feeding MRR & 116 \\
		MRR time-series & 232 \\
		\bottomrule
	\end{tabular}%
	\caption{\textbf{Summary of variables across all MRR time-series.} For the `Species' and `Genera' rows, the counts represent the number of distinct entries in the database; otherwise the counts represent the number of time series.}
	\label{tab:mrr_aggregateData}%
\end{table}%

\begin{table}[htbp]
	\centering
	\footnotesize
	\begin{adjustwidth}{-0.5in}{-0.5in}%
		\begin{tabularx}{1.25\textwidth}{l|ccccc}
			\toprule
			\textbf{Data characteristic} & \textbf{Min} & \textbf{Mean} & \textbf{Median} & \textbf{Max} & \textbf{Standard deviation} \\
			\midrule
			Study duration, days & 6.0   & 11.8  & 10.0  & 71.0    & 4.7 \\
			Number of days on which collections took place & 6.0   & 10.1  & 9.0   & 47    & 9.1 \\
			Number of separate release days & 1.0     & 1.9   & 1.0     & 23    & 3.0 \\
			Number released & 66    & 4,929  & 1,297  & 86,200 & 12,043 \\
			Number recaptured & 2     & 163   & 63    & 4,090  & 399 \\
			Recapture percentage & 0\% & 8.6\% & 5.2\% & 57.1\% & 10.04\% \\
			Age at release, days & 0   & 1.8   & 0   & 13    & 2.8 \\
			\bottomrule
		\end{tabularx}%
		\caption{\textbf{Summary of data from individual MRR time-series.}}
		\label{tab:mrr_IndividualData}%
	\end{adjustwidth}
\end{table}%

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.25\textwidth]{./Figure_files/mrr_exampleMRRSeries.pdf}}
	\caption{\textbf{Example MRR time-series for three different studies from the \cite{guerra2014global} database.} In all the above experiments there was a single release of a batch of marked mosquitoes.}
	\label{fig:mrr_exampleMRRSeries}
\end{figure}

The simplest assumption is that death and dispersal rates are independent of mosquito age, giving an exponential `survival' function,
%
\begin{equation}
S(t) = e^{-\lambda t},
\end{equation}
%
where $\lambda$ is the sum of the rates of death and dispersal from the study area. To estimate the lifespan of mosquitoes we use this model, partly because of the limited evidence we determine in favour of senescence (Figs. \ref{fig:mrr_elpd} and \ref{fig:dissection_elpd}).

@@

\subsection{Individual time-series estimates}\label{sec:MRR_individual_analysis}
We first treat each time-series separately, and estimate individual $(\lambda,\psi,\kappa)$ parameters of the statistical model (described in the previous section) for each time-series. To use a Bayesian methodology we must specify priors on this set of parameters. Here we choose to specify independent priors on each parameter of the form: $\lambda\sim \mathcal{N}(-2.32,1)$, $\psi\sim \text{exp}(50)$ (with $\psi$ constrained to be less than 0.1, a threshold below which the majority of daily recapture probabilities in the data are observed) and $\kappa\sim \textnormal{log-normal}(2,1)$. This prior on the rate parameter of the exponential distribution ($\lambda$) corresponds to a wide range of possible lifespans (Fig. \ref{fig:mrr_individualTimeSeries_priors}A), with a mean of 10 days. It was necessary to use an informative prior on $\psi$ to bound it to sensible values and to prevent the correlation of estimates with $\lambda$ leading to unreasonably short estimates of $\lambda$; the prior on $\kappa$ is fairly uninformative and allows a wide range of values (Fig. \ref{fig:mrr_individualTimeSeries_priors}B,C).

@@

\subsection{Estimating lifespan at the species, genus and overall groupings}\label{sec:MRR_hierarchical}
We synthesise information from across all MRR experiments to produce more robust estimates of mosquito lifespan than can be obtained from considering the individual time-series separately. However there exists considerable heterogeneity across the experiments. This heterogeneity has two sources. There is that arising from variability in experimental methodology. There is also variability from actual differences in lifespan across the different mosquito cohorts; for example due to genetic differences between mosquito populations or due to climatic differences. Because of this heterogeneity we use a Bayesian hierarchical model which is akin to a `random effects' model in classical statistics. This type of model assumes that there is random variation in parameters at the individual time-series level, although each of the parameters is drawn from a common `population-level' distribution. In our case we separately estimate three different model groupings where the population-level distributions correspond to the species, genus and overall (across all studies) levels respectively (Fig. \ref{fig:mrr_methodDiagramHierarchy}). This allows us to estimate average mosquito lifespan for a given species or genus, or alternatively across all time-series, by independent sampling from the posterior predictive distribution from which the individual $\lambda_i$ are drawn.

@@

Whilst we assume that the variation in mosquito mortality across experiments is random in nature, we recognise that there may exist systematic sources of variation that we do not include in our model. To account for this potential source of bias we examined how the individual estimates of lifespan correlated with experimental covariates (trapping method and number or density of traps). The data on trapping method in the Guerra et al., (2014) database is less reliable than other data entries because published MRR studies may use multiple trapping methods and typically do not disaggregate captures by method. When we examined the correlation of trapping method with lifespan estimates, however, we did not find systematic variation by method (data not shown). We determined a positive correlation between trapping density and lifespan estimates (Fig. \ref{fig:mrr_lifeSpanVsTrapDensity}), which indicates the impact of mosquito dispersal on estimated lifespan.

A Bayesian framework requires that we incorporate our pre-analysis beliefs into our estimates through the use of priors for all parameters in a particular model. In our hierarchical model we are required to specify prior distributions at two levels of the analysis. The first of these links the individual time-series estimates with the overarching group-level distributions. The parameters of these group-level distributions are then set prior distributions (Fig. \ref{fig:mrr_methodDiagramHierarchy}). As an example, we suppose that the rate parameter $\lambda_i$ of our exponential survival model in experiment $i$ is given by:
%
\begin{align}
\lambda_i = \text{exp}(&c_i + \delta^m_{\text{genus}[i]} \text{sex}^{\text{m}}_i + \delta^{\text{mix}}_{\text{genus}[i]} \text{sex}^{\text{mix}}_i +\\  & \delta^{\text{sugar}}_{\text{genus}[i]} \text{sugar}_i +  \delta^{\text{blood}}_{\text{genus}[i]} \text{blood}_i),
\end{align}
%
where $\delta^m_{\text{genus}[i]}$ measures a genus-specific effect of male-only releases ($\text{sex}^{\text{m}}_i=1$) versus female-only ($\text{sex}^{\text{m}}_i=0$); $\delta^{\text{mix}}_{\text{genus}[i]}$ is the analogous effect for mixed releases versus female-only; $\delta^{\text{sugar}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were sugar-fed prior to release versus those that were not fed; and $\delta^{\text{blood}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were blood-fed prior to release versus those that were not fed. The parameter $c_i$ is modelled hierarchically, being drawn from a group-level distribution assumed to be normal,
%
\begin{equation}
c_i \sim  \mathcal{N}(\mu_\lambda,\sigma)
\end{equation}
%
We then specify priors on these group-level parameters, $\mu_\lambda \sim \mathcal{N}(-2.32,1)$ and $\sigma \sim \textnormal{log-normal}(-3,1)$. On all $\delta$ parameters, except $\delta^{\text{mix}}_{\text{genus}[i]}$, we specify standard normal priors. We specify a uniform prior on $\delta^{\text{mix}}_{\text{genus}[i]}\sim U(0,\delta^m_{\text{genus}[i]})$ so that the effect of mixed releases lies between the female-only release effect (0) and the male-only release effect ($\delta^m_{\text{genus}[i]}$).

The relatively complex nature of priors in hierarchical models make it important to determine their influence on inferences. We chose the above -- somewhat uninformative -- priors to allow a range of mosquito lifespans in order to minimise their effect on the estimates we report (Fig. \ref{fig:mrr_meanLifespanPrior}). We also chose to set hierarchical priors on the remaining parameters in our models -- $\psi$ the probability of daily recapture, and $\kappa$ the `over-dispersion' parameter of a negative binomial distribution. For $\psi$ we chose an informative prior that placed all probability mass $\psi\leq 10\%$, since in most experiments the fraction of mosquitoes captured was significantly below this value. For $\kappa$ we chose a fairly wide prior that had most of its probability mass below $\kappa=20$ (Fig. \ref{fig:mrr_PsiKappaPriors}). In all cases the hierarchical priors were chosen to have similar implications on lifespan, recapture probability and over-dispersion at the individual time-series level as for the non-hierarchical analysis described previously.

\begin{figure}[h]
	\centerline{\includegraphics[width=1\textwidth]{./Figure_files/mrr_lifespan_priors.pdf}}
	\caption{\textbf{Samples from the prior predictive distribution of mean lifespan across the six models of mosquito mortality introduced in Table \ref{tab:mrr_survivalDescription}.} In all cases 10,000 samples were generated using the priors indicated in Table \ref{tab:mrr_priors}, resulting in distributions with a mean close to 10 days.}
	\label{fig:mrr_meanLifespanPrior}
\end{figure}

@@

\subsection{Testing for age-dependent mortality in wild mosquitoes}\label{sec:mrr_age_dependence}
Previous work has found evidence of age-dependent mortality in lab populations \citep{styer2007mosquitoes,dawes2009anopheles}, and less-commonly in wild mosquitoes \citep{clements1981analysis}. Furthermore, recent modelling work has examined the implications of departures from a constant risk of mortality \citep{styer2007mosquitoes,hancock2009age,novoseltsev2012age}. To determine whether senescence occurs in wild mosquitoes we re-estimate our model using survival functions, $S(t)$, that allow for a rate of mortality that can vary with age. Specifically, we re-estimate our model using five other models that were previously used in the literature (see Table \ref{tab:mrr_survivalDescription} for a description of these). Each of these models makes different assumptions about how the rate of death and dispersal varies with age, but all can be represented by a general form,
%
\begin{equation}\label{eq:survivalInt}
S(t) = e^{-\int\limits_{0}^{t}\lambda(\tau) \mathrm{d}\tau}.
\end{equation}
%
where we constrain the parameters of our models to preclude the possibility of a hazard that decreases with age ($\frac{\mathrm{d}\lambda(t)}{\mathrm{d}t} \geq 0$). Whilst this could occur if older mosquitoes disperse less than younger mosquitoes, it is unlikely this would outweigh any declines in survival associated with old age.

The parameters of the survival function of each model were assigned hierarchical priors that allowed considerable variation in mosquito lifespan (see Fig. \ref{fig:mrr_meanLifespanPrior}). Otherwise the statistical model we used remained the same as for the constant mortality case. In testing for age-dependent mortality, we did not account for sex or pre-release feeding effects due to the complexity of including these elements in more complex survival models.

\begin{table}
	\scriptsize
	\noindent\makebox[\textwidth]{%
		\begin{tabularx}{1.35\textwidth}{c|cXcX}
			\textbf{Survival function} & \textbf{Hazard rate} & \textbf{Interpretation} & \textbf{Mean time spent in study area, }\textbf{$\overline{T}$} & \textbf{Papers assuming this function} \\
			\midrule
			Exponential & $\lambda$ & Constant mortality risk & $\frac{1}{\lambda}$ & \cite{ross1910prevention}; \cite{anderson1992infectious}; \cite{smith2004statics} \\
			\midrule
			Gompertz &  $\alpha e^{\beta t}$     & Mortality increases with age at an ever-increasing rate &  $\frac{e^{\frac{\alpha}{\beta}} \Gamma(0,\frac{\alpha}{\beta})}{\beta}$    & \cite{clements1981analysis}; \cite{styer2007mosquitoes};\cite{novoseltsev2012age} \\
			\midrule
			Weibull &  $\alpha t^\beta$; $\;\beta\geq 1$     &  Mortality increases with age at an ever-increasing rate     &   $\alpha^{-\frac{1}{\beta}} \Gamma(1+\frac{1}{\beta})$    & \cite{hancock2009age}; and \cite{carey2001insect} considering general insect demography \\
			\midrule
			Gompertz-Makeham &  $\alpha e^{\beta t} + c$      &  Two additive mortality risks: one that increases with age, and another that is age-independent  &  No simple analytic form     & \cite{styer2007mosquitoes} \\
			\midrule
			Logistic & $\frac{\alpha e^{\beta t}}{1+\frac{\alpha s}{b}\left(e^{b t} - 1\right)}$   & Mortality risk increases with age, although at a declining rate  & No simple analytic form   & \cite{styer2007mosquitoes}; \cite{novoseltsev2012age} \\
			\midrule
			Logistic-Makeham &   $\frac{\alpha e^{\beta t}}{1+\frac{\alpha s}{b}\left(e^{b t} - 1\right)} + c$    & Two separate additive mortality risks: an age-dependent risk of the same form as the logistic model, and a constant hazard  &  No simple analytic form  & \cite{styer2007mosquitoes}; \cite{bellan2010importance} \\
			\bottomrule
	\end{tabularx}}
	\caption{\textbf{A description of the survival functions used in this study, arranged in rough order of model complexity (simple-complex from top-bottom.)} All parameters are defined to be non-negative. $\Gamma(\theta)$ and $\Gamma(\theta_1,\theta_2)$ refer to the Euler gamma function, and incomplete gamma function respectively. The `mean time spent in study area' is an estimate of the combined effects of mosquito mortality and dispersal from the area of the study where collections take place, since our data does not provide spatially-resolved data.}\label{tab:mrr_survivalDescription}
\end{table}

>>>>

\begin{itemize}
	\item 69 and 70: Since the release was not disaggregated by sex, the recapture data were aggregated across both males and females, resulting in a single time series in each case.
	\item 482: There are actually two series (one with captured mosquitoes, the other with young reared) in the paper \cite[]{rawlings1981influence}, whereas the database contains only one recapture series. In this analysis, both time series are included.
	\item 483: There are three studies in the paper \cite[]{rawlings1981influence}, (by dye colour: there is one for `M\&Y', another for white, and a third for green), whereas the database only contains one release series. In this analysis, we analyse all three time series separately.
	\item 578: A null observation was removed on day 12, since it isn't present in the original paper \cite[]{eyles1943experiment}.
	\item 585: It is not clear from the original paper \cite[]{smith1941observations} whether missing recapture observations on certain days are because no mosquitoes were found on those days, or because no efforts were made to recapture them on those days. We have assumed that the number recaptured were as stated in the database on those days (missing, because no recapture efforts were made on those days).
	\item 586: It was unclear from database whether the missing observations are due to lack of recapture, or no effort. Have assumed that zero mosquitoes were captured on these days (the paper could not be accessed).
	\item 342: Changed number of released from 6,000 to 12,000 following the original publication \cite[]{bryan1991anopheles}.
	\item 246: Moved the observations from captures to recaptures following the original publication \cite[]{arredondo1998gonotrophic}.
	\item 55: Corrected a mistake in release data; the actual number released is 739, not 749 \cite[]{midega2007estimating}.
	\item 478-481: All four of these series give releases not disaggregated by gender, whereas the recaptures are split this way. Have combined the recapture series to yield a single release-recapture series in each case.
	\item 154: Moved captured male series to recaptured following original paper \cite[]{tsuda2001movement}. 
	\item 334: Data starts on day 3, thus have added this number of days to our time variable in this case.
\end{itemize}

The R file \verb|s_clean_data.R| cleans the data, makes the above amendments and outputs the data in a form amenable for estimation using Stan \citep{carpenter2016stan}.

@@

\begin{table}[htbp]
	\centering
	\begin{tabular}{lc}
		\toprule
		\textbf{Data characteristic} & \textbf{Count} \\
		\midrule
		Species & 33 \\
		Genera & 3 \\
		\textit{Anopheles} & 94 \\
		\textit{Aedes} & 91 \\
		\textit{Culex} & 47 \\
		Female-only MRR & 179 \\
		Male-only MRR & 35 \\
		Mixed sex MRR & 18 \\
		Pre-release blood-feeding only MRR & 71 \\
		Pre-release sugar-feeding only MRR & 41 \\
		Pre-release both blood- and sugar-feeding MRR & 4 \\
		Pre-release neither blood- and sugar-feeding MRR & 116 \\
		MRR time-series & 232 \\
		\bottomrule
	\end{tabular}%
	\caption{\textbf{Summary of variables across all MRR time-series used in hierarchical analyses.} For the `Species' and `Genera' rows, the counts represent the number of distinct entries in the database; otherwise the counts represent the number of time series.}
	\label{tab:mrr_aggregateData}%
\end{table}%

\begin{table}[htbp]
	\centering
	\footnotesize
	\begin{adjustwidth}{-0.5in}{-0.5in}%
		\begin{tabularx}{1.25\textwidth}{l|ccccc}
			\toprule
			\textbf{Data characteristic} & \textbf{Min} & \textbf{Mean} & \textbf{Median} & \textbf{Max} & \textbf{Standard deviation} \\
			\midrule
			Study duration, days & 6.0   & 11.8  & 10.0  & 71.0    & 4.7 \\
			Number of days on which collections took place & 6.0   & 10.1  & 9.0   & 47    & 9.1 \\
			Number of separate release days & 1.0     & 1.9   & 1.0     & 23    & 3.0 \\
			Number released & 66    & 4,929  & 1,297  & 86,200 & 12,043 \\
			Number recaptured & 2     & 163   & 63    & 4,090  & 399 \\
			Recapture percentage & 0\% & 8.6\% & 5.2\% & 57.1\% & 10.04\% \\
			Age at release, days & 0   & 1.8   & 0   & 13    & 2.8 \\
			\bottomrule
		\end{tabularx}%
		\caption{\textbf{Summary of data from individual MRR time-series used in hierarchical analyses.}}
		\label{tab:mrr_IndividualData}%
	\end{adjustwidth}
\end{table}%

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.25\textwidth]{./Figure_files/mrr_exampleMRRSeries.pdf}}
	\caption{\textbf{Example MRR time-series for three different studies from the \cite{guerra2014global} database.} In all the above experiments, there was a single release of marked mosquitoes on day zero.}
	\label{fig:mrr_exampleMRRSeries}
\end{figure}

The simplest assumption is that death and dispersal rates are independent of mosquito age, giving an exponential `survival' function,
%
\begin{equation}
S(t) = e^{-\lambda t},
\end{equation}
%
where $\lambda$ is the sum of the rates of death and dispersal from the study area. To estimate the lifespan of mosquitoes we use this model, partly because of the limited evidence we find for senescence.

@@

\subsection{Individual time-series estimates}\label{sec:MRR_individual_analysis}
We first treat each time-series separately, and estimate individual $(\lambda,\psi,\kappa)$ parameters of the statistical model (described in Section \ref{sec:mrr_statistical}) for each time-series. To use a Bayesian methodology, we must specify priors on this set of parameters. Here, we choose to specify independent priors on each parameter of the form: $\lambda\sim \mathcal{N}(-2.32,1)$, $\psi\sim \text{exp}(50)$ (with $\psi$ constrained to be less than 0.1, a threshold below which the majority of daily recapture probabilities in the data are observed) and $\kappa\sim \textnormal{log-normal}(2,1)$. This prior on the rate parameter of the exponential distribution ($\lambda$) corresponds to a wide range of possible lifespans (Fig. \ref{fig:mrr_individualTimeSeries_priors}A), with a mean of 10 days. It was necessary to use an informative prior on $\psi$ to bound it to sensible values and to prevent unreasonably short estimates of $\lambda$; the prior on $\kappa$ is fairly uninformative and allows a wide range of values (Fig. \ref{fig:mrr_individualTimeSeries_priors}B,C).

@@

\subsection{Estimating lifespan at the species, genus and overall groupings}\label{sec:MRR_hierarchical}
We synthesise information from across all MRR experiments to produce more robust estimates of mosquito lifespan than can be obtained from considering the individual time-series separately. However, there exists considerable heterogeneity across the experiments. This heterogeneity has two sources: that arising from variability in experimental methodology; but also that from actual differences in lifespan across the different mosquito cohorts -- for example, due to genetic differences between mosquito populations or due to climatic differences. Because of this heterogeneity, we use a Bayesian hierarchical model which is akin to a `random effects' model in classical statistics. This type of model assumes that there is random variation in parameters at the individual time-series level, although each of the parameters is drawn from a common `population-level' distribution. In our case, we separately estimate three different model groupings where the population-level distributions correspond to the species, genus and overall (across all studies) levels respectively (Fig. \ref{fig:mrr_methodDiagramHierarchy}). This allows us to estimate mosquito lifespan for a given species or genus by independent sampling from the posterior predictive distribution from which the individual $\lambda_i$ are drawn.

@@

Whilst we assume that the variation in mosquito mortality across experiments is random in nature, we recognise that there may exist systematic sources of variation that we do not include in our model. To account for this potential source of bias, we examined how the individual estimates of lifespan correlated with experimental covariates (number and density of traps). From this, we determined a slight but insignificant positive correlation between trapping number and density with individual lifespan estimates -- suggesting that downward bias due to insufficient trapping may be relatively small.

In our hierarchical model, we must specify prior distributions at two levels of the analysis. The first of these links the individual time-series estimates with the overarching group-level distributions. The parameters of these group-level distributions are then set prior distributions (Fig. \ref{fig:mrr_methodDiagramHierarchy}). As an example, suppose that the rate parameter $\lambda_i$ of our exponential survival model in experiment $i$ is given by:
%
\begin{align}
\lambda_i = \text{exp}(&c_i + \delta^m_{\text{genus}[i]} \text{sex}^{\text{m}}_i + \delta^{\text{mix}}_{\text{genus}[i]} \text{sex}^{\text{mix}}_i +\\  & \delta^{\text{sugar}}_{\text{genus}[i]} \text{sugar}_i +  \delta^{\text{blood}}_{\text{genus}[i]} \text{blood}_i),
\end{align}
%
where $\delta^m_{\text{genus}[i]}$ measures a genus-specific effect of male-only releases ($\text{sex}^{\text{m}}_i=1$) versus female-only ($\text{sex}^{\text{m}}_i=0$); $\delta^{\text{mix}}_{\text{genus}[i]}$ is the analogous effect for mixed releases versus female-only; $\delta^{\text{sugar}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were sugar-fed prior to release versus those that were not fed; and $\delta^{\text{blood}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were blood-fed prior to release versus those that were not fed. The parameter $c_i$ is modelled hierarchically, being drawn from a group-level distribution assumed to be normal,
%
\begin{equation}
c_i \sim  \mathcal{N}(\mu_\lambda,\sigma).
\end{equation}
%
We then specify priors on these group-level parameters, $\mu_\lambda \sim \mathcal{N}(-2.32,1)$ and $\sigma \sim \textnormal{log-normal}(-3,1)$. On all $\delta$ parameters, except $\delta^{\text{mix}}_{\text{genus}[i]}$, we specify standard normal priors. We specify a uniform prior on $\delta^{\text{mix}}_{\text{genus}[i]}\sim U(0,\delta^m_{\text{genus}[i]})$, so that the effect of mixed releases lies between the female-only release effect (0) and the male-only release effect ($\delta^m_{\text{genus}[i]}$).

The relatively complex nature of priors in hierarchical models make it important to determine their influence on inferences. We chose the above -- somewhat uninformative -- priors, to allow a range of mosquito lifespans in order to minimise their effect on the estimates we report (Fig. \ref{fig:mrr_meanLifespanPrior}). We also chose to set hierarchical priors on the remaining parameters in our models -- $\psi$ the probability of daily recapture, and $\kappa$ the `over-dispersion' parameter of a negative binomial distribution. For $\psi$, we chose an informative prior that placed all probability mass $\psi\leq 10\%$, since in most experiments the fraction of mosquitoes captured was significantly below this value. For $\kappa$, we chose a fairly wide prior that had most of its probability mass below $\kappa=20$ (Fig. \ref{fig:mrr_PsiKappaPriors}). In all cases, the hierarchical priors were chosen to have similar implications on lifespan, recapture probability and over-dispersion at the individual time-series level as for the non-hierarchical analysis described previously.

\begin{figure}[h]
	\centerline{\includegraphics[width=1\textwidth]{./Figure_files/mrr_lifespan_priors.pdf}}
	\caption{\textbf{Samples from the prior predictive distribution of mean lifespan across the six models of mosquito mortality introduced in Table \ref{tab:mrr_survivalDescription}.} In all cases, 10,000 samples were generated using the priors indicated in Table \ref{tab:mrr_priors}, resulting in distributions with a mean close to 10 days.}
	\label{fig:mrr_meanLifespanPrior}
\end{figure}

@@

\subsection{Testing for age-dependent mortality in wild mosquitoes}\label{sec:mrr_age_dependence}
Previous work has found evidence of age-dependent mortality in lab populations \citep{styer2007mosquitoes,dawes2009anopheles}, and less-commonly in wild mosquitoes \citep{clements1981analysis,harrington2008age}. Furthermore, modelling work has examined the implications of departures from a constant risk of mortality \citep{styer2007mosquitoes,hancock2009age,novoseltsev2012age}. To determine whether senescence occurs in wild mosquitoes, we re-estimate the model using a range of different survival functions, $S(t)$, that permit mortality to increases with age. Specifically, we re-estimate our model using five other models that were previously used in the literature (see Table \ref{tab:mrr_survivalDescription} for a description of these). Each of these models makes different assumptions about how the rate of death and dispersal varies with age, but all can be represented by a general form,
%
\begin{equation}\label{eq:survivalInt}
S(t) = e^{-\int\limits_{0}^{t}\lambda(\tau) \mathrm{d}\tau}.
\end{equation}
%
where we constrain the parameters of our models to preclude the possibility of a hazard that decreases with age ($\frac{\mathrm{d}\lambda(t)}{\mathrm{d}t} \geq 0$). Whilst this could occur if older mosquitoes disperse less than younger mosquitoes, it is unlikely this would outweigh any declines in survival associated with old age.

The parameters of the survival function of each model were assigned hierarchical priors that allowed considerable variation in mosquito lifespan (see Fig. \ref{fig:mrr_meanLifespanPrior}) but had similar means as to the constant mortality case. Otherwise, the statistical model we used remained the same as for the constant mortality case. In testing for age-dependent mortality, we did not account for sex or pre-release feeding effects due to the complexity of including these elements in more complex survival models.

\begin{table}
	\scriptsize
	\noindent\makebox[\textwidth]{%
		\begin{tabularx}{1.35\textwidth}{c|cXcX}
			\textbf{Survival function} & \textbf{Hazard rate} & \textbf{Interpretation} & \textbf{Mean time spent in study area, }\textbf{$\overline{T}$} & \textbf{Papers assuming this function} \\
			\midrule
			Exponential & $\lambda$ & Constant mortality risk & $\frac{1}{\lambda}$ & \cite{ross1910prevention}; \cite{anderson1992infectious}; \cite{smith2004statics} \\
			\midrule
			Gompertz &  $\alpha e^{\beta t}$     & Mortality increases with age at an ever-increasing rate &  $\frac{e^{\frac{\alpha}{\beta}} \Gamma(0,\frac{\alpha}{\beta})}{\beta}$    & \cite{clements1981analysis}; \cite{styer2007mosquitoes};\cite{novoseltsev2012age} \\
			\midrule
			Weibull &  $\alpha t^\beta$; $\;\beta\geq 1$     &  Mortality increases with age at an ever-increasing rate     &   $\alpha^{-\frac{1}{\beta}} \Gamma(1+\frac{1}{\beta})$    & \cite{hancock2009age}; and \cite{carey2001insect} considering general insect demography \\
			\midrule
			Gompertz-Makeham &  $\alpha e^{\beta t} + c$      &  Two additive mortality risks: one that increases with age, and another that is age-independent  &  No simple analytic form     & \cite{styer2007mosquitoes} \\
			\midrule
			Logistic & $\frac{\alpha e^{\beta t}}{1+\frac{\alpha s}{b}\left(e^{b t} - 1\right)}$   & Mortality risk increases with age, although at a declining rate  & No simple analytic form   & \cite{styer2007mosquitoes}; \cite{novoseltsev2012age} \\
			\midrule
			Logistic-Makeham &   $\frac{\alpha e^{\beta t}}{1+\frac{\alpha s}{b}\left(e^{b t} - 1\right)} + c$    & Two separate additive mortality risks: an age-dependent risk of the same form as the logistic model, and a constant hazard  &  No simple analytic form  & \cite{styer2007mosquitoes}; \cite{bellan2010importance} \\
			\bottomrule
	\end{tabularx}}
	\caption{\textbf{A description of the survival functions used in this study, arranged in rough order of model complexity (simple-complex from top-bottom.)} All parameters are defined to be non-negative. $\Gamma(\theta)$ and $\Gamma(\theta_1,\theta_2)$ refer to the Euler gamma function, and incomplete gamma function respectively. The `mean time spent in study area' is an estimate of the combined effects of mosquito mortality and dispersal from the area of the study where collections take place, since we do not analyse spatially-resolved data.}\label{tab:mrr_survivalDescription}
\end{table}

\begin{table}
	\footnotesize
	\noindent\makebox[\textwidth]{%
		\begin{tabularx}{1.2\textwidth}{l|c|X}
			\textbf{Survival function} & \textbf{time-series-level priors} & \textbf{Group-level priors}\\
			\midrule
			Exponential & $c\sim \mathcal{N}(\mu_\lambda,\sigma)$ &$\mu_\lambda \sim N(-2.32,1)\text{,}\; \sigma\sim \textnormal{log-normal}(-3,1)$\\
			\midrule
			Gompertz & $\alpha,\beta\sim$ log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,1.1)$\\
			\midrule
			Weibull & $\alpha,(\beta-1)\sim$  log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-4.8,1.75)\text{,}\; \mu_\beta\sim N(-4,2)$\\
			\midrule
			Gompertz-Makeham & $\alpha,\beta,c\sim$ log-normal$(\mu_{\alpha|\beta|c},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4.5,0.5)$\\
			\midrule
			Logistic & $\alpha,\beta,s\sim$  log-normal$(\mu_{\alpha|\beta|s},0.2)$ &$\mu_\alpha\sim N(-2.8,1)\text{,}\; \mu_\beta\sim N(-3,1)\text{,}\; \mu_s\sim N(-3,1)$\\
			\midrule
			Logistic-Makeham & $\alpha,\beta,s,c\sim$ log-normal$(\mu_{\alpha|\beta|s|c},0.2)$ &$\mu_\alpha\sim N(-3.2,1)\text{,}\; \mu_\beta\sim N(-3.2,1)\text{,}\; \mu_s\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4,0.5)$\\
			\bottomrule
	\end{tabularx}}\caption{\textbf{The priors used on parameters of each different survival model.} For the exponential model, the `group-level' priors were the same for the genus and `overall' models that were also estimated. The exponential model was the only model that was simple enough to allow the scale parameter of the log-normal ($\sigma$) to be estimated by the data. The notation $\alpha,\beta\sim \text{log-normal}(\mu_{\alpha|\beta},0.2)$ means that $\alpha$ and $\beta$ were assigned independent log-normal priors with location parameters $\mu_\alpha$ and $\mu_\beta$ respectively, and a scale parameter of 0.2 in both cases.}\label{tab:mrr_priors}
\end{table}

\subsection{Model estimation by MCMC}\label{sec:mrr_MCMC}
The likelihood and priors used result in posterior distributions whose analytic form cannot be calculated with existent computational methods. Instead, we use Markov Chain Monte Carlo (MCMC) methods to sample from each posterior distribution. Specifically, we used \textit{Stan} software \citep{carpenter2016stan} that implements an efficient MCMC algorithm known as NUTS \citep{hoffman2014no}.

To judge convergence of the sampling algorithm to the posterior density, we calculated $\hat{R}$ across all Markov chains -- a ratio that compares between-chain variation to that within each chain that is commonly used to measure convergence in MCMC \citep{gelman1992inference}. For each class of model, we used the following MCMC parameters,
%
\begin{itemize}
	\item Individual lifespan models (Section \ref{sec:MRR_individual_analysis}): 12 independent chains with 1000 iterations per chain,
	\item Hierarchical lifespan estimates (Section \ref{sec:MRR_hierarchical}): 12 independent chains with 15,000 iterations per chain,
	\item Age-dependent analysis (Section \ref{sec:mrr_age_dependence}): 12 independent chains with 3000 iterations per chain.
\end{itemize}

In all cases, we discarded the first half of these iterations as warm-up \citep{gelman2014bayesian}. At the end of all runs for lifespan estimation, $\hat{R}<1.1$ for all model parameters. For the age-dependent cases, there remained a handful of parameters where $\hat{R}>1.1$ after running simulations for 3000 iterations but this did not affect our inferences, since repeated runs demonstrated the same pattern. We also ensured that across each MCMC run, the number of divergent iterations (that can bias the MCMC away from the true posterior density) was minimal. In the majority of cases, the number of divergent iterations was far fewer than 1\% of the total number of samples.

The Stan scripts for lifespan estimation (i.e. those using an exponential survival model) for the MRR analyses are provided below. The script for the individual time series analysis (Section \ref{sec:MRR_individual_analysis}):
